Anxiety, avoidance, and sequential evaluation

### 2019-02-28 manuscripts/rldm2019/rldm2019.tex
further rldm2019 updates

\begin{abstract}
Anxiety disorders are characterized by aberrations in the processing and response to threat, involving symptoms like exaggerated threat appraisal, fear generalization, and persistent avoidance. These symptoms are puzzling from a decision theoretic perspective. The generalization of fear and avoidance to situations far antecedent of threat violates the core logic of evaluation over sequential trajectories of action. This is because avoidance is by nature protective: hence, the ability to avoid in future means I need not also do so now. This is an instance of a general property of evaluating actions in sequential tasks: such evaluation turns fundamentally on assumptions about subsequent events, including the agent's own subsequent choices. Typically, it is appropriate to assume that the agent will also make good choices down the line. This line of reasoning, then, hints that a core deficit in anxious disorders may relate to this assumption, which otherwise should preclude the spread of avoidance. Indeed, anxiety disorders are associated with pessimistic beliefs about the future. Here we develop this idea into a simple model of evaluation under pessimistic assumptions about future choices. We show that this single, minor deviation can explain a surprising range of features of anxious behavior  including exaggerated threat appraisals, fear generalization, and persistent avoidance. This model also offers a new formalization of classic insights from the psychiatric literature about the central role of beliefs about control and self-efficacy in anxiety. This perspective is importantly different from previous computational accounts of beliefs about control in mood disorders, which neglected the sequential aspects of choice. 
\end{abstract}

@@

\acknowledgements{lalalalala.}

@@

These symptoms are particularly puzzling from a decision theoretic perspective \citep{huys2015}. Here we identify one anomaly, which we suggest offers fundamental insights into the core of the dysfunction: The generalization of fear and avoidance to situations far antecedent of threat violates the core logic of evaluation over sequential trajectories of action. This is because avoidance is by nature protective: the ability to avoid in future means I need not also do so now. For instance, cars endanger pedestrians but can be avoided by following traffic signals; given that, staying indoors offers no additional protection from accidents. This is an instance of a general property of evaluating actions in sequential tasks: such evaluation turns fundamentally on assumptions about subsequent events, including the agent's own subsequent choices. Typically, it is appropriate to assume that the agent will also make good choices down the line.

This line of reasoning, then, hints that a core deficit in anxiety disorders may relate to this assumption, which otherwise should preclude the spread of avoidance. Indeed, anxiety disorders are associated with pessimistic beliefs about the future \citep{ClarkBeck2011}. For instance, anxious individuals are more likely to endorse beliefs like that they are likely to encounter danger in the future and that they will be unable to mitigate it when they do. Here we develop this idea into a simple model of evaluation under pessimistic assumptions about future choices. We show that this single, minor deviation can explain a surprising range of features of anxious behavior including exaggerated threat appraisal, fear generalization, and persistent avoidance. This model also offers a new formalization of classic insights from the psychiatric literature about the central role of beliefs about control and self-efficacy in anxiety. This perspective is importantly different from previous computational accounts of beliefs about control in mood disorders (e.g. \cite{HuysDayan2009}), which neglected the sequential aspects of choice.

@@

Consider an agent who instead has pessimistic expectations about their future actions. Such returns can be adaptive to achieve robustness such as under adversarial scenarios \cite{Garcia2015}; here we propose unrealistically pessimistic assumptions as a root cause of many anxious symptoms. We adopt the $\beta$-pessimistic value function from \cite{Gaskett2003}, to define
\begin{equation*}
Q^w(s,a) = r(s,a) + \gamma \sum_{s'} p(s' \mid s,a) \left( w \max_{a'} Q^w(s',a') + (1 - w) \min_{a'} Q^w(s',a') \right)
\end{equation*}
Here, the parameter $w$ controls the degree of pessimism. An optimistic agent ($w = 1$) expects in the future to act fully in accordance with its preferences whereas a pessimistic agent ($w = 0$) expects to act contrary to its preferences. Fig.~1c,d illustrates the the consequences of different levels of pessimism for valuation in the example gridworld; here, the expectation that the agent may fail to avoid in future causes the value of threat to permeate the state space.

@@

This simple simulation --- reflecting a small violation of a core decision theoretic assumption --- already echoes several core symptoms of anxiety disorders. The pessimistic agents in Fig~1c,d exhibit exaggerated threat appraisals (otherwise neutral states signal danger); generalization of fear (threat value spreads across the gridworld); and persistent avoidance (the agent chooses paths which maintain increasing distance from threat).

@@

\textbf{Aversive pruning:} Another empirical phenomenon associated with anxiety is ``aversive pruning'' in planning \cite{Huys2012, Lally2017}. This refers to the idea that when evaluating future trajectories in a sequential task like chess, people must selectively consider certain paths and neglect others. One proposal for how people accomplish this is aversive pruning \citep{Huys2012}: choice sequences involving large losses are discarded from further evaluation. For instance, although the optimal choice in the decision tree from Fig.~3a is to weather an initial large loss in order to reap the large gain that follows, people tend to disfavor this path suggesting they neglect the later gain. The degree of such pruning correlates with depressive \cite{Huys2012} and anxiety \cite{Lally2017} symptoms. Our model predicts this result (Fig.3~b), though for a somewhat different reason: in our model, pessimistic (anxious) agents neglect large gains deeper in the tree, not because they fail to consider them, but because with increasing anxiety they increasingly predict the possibility of choosing incorrectly afterward, and failing to recoup the loss. Future research could use slight variants in the decision trees could be used to tease apart these different interpretations.

@@

@@

One important ambiguity is that we have formalized pessimistic assumptions in terms of beliefs about the agent's own future actions: expecting failure, for instance, to avoid in future. Since Eq.~1 is also computed in expectation over the anticipated future environmental dynamics $p(s' \mid s,a)$, pessimism can alternatively be encoded in this distribution (e.g., a belief that the world's response to one's choices is unpredictable or adversarial). Since the Bellman equation averages out both distributions, either formulation can produce ultimately similar results. However from the perspective of cognitive theories of anxiety, these represent quite different maladaptive beliefs, which may be relevant, for instance, in guiding treatment.

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/04_tree_rldm.png}}%
  }
  \par \textbf{Figure 3:} The decision tree environment from \cite{Huys2012} and \cite{Lally2017} (A). An optimistic agent ($w=1$) prefers the optimal loss-minimizing policy through the initial large loss (B). A pessimistic agent ($w=0.5$) comes to prefer the branch without the large loss so as to avoid being unable to recoup the large initial loss. The net value $Q$ on each path is shown numerically.
\end{figure}

>>>>

\begin{abstract}
Anxiety disorders are characterized by a range of aberrations in the processing and response to threat, but there is little clarity what core pathogenesis might underlie these symptoms. Here we propose a decision theoretic analysis of maladaptive avoidance and embody it in a reinforcement learning model, which shows how a localized bias in beliefs can formally explain a range of phenomena related to anxiety. The core observation, implicit in standard decision theoretic accounts of sequential evaluation, is that avoidance should be protective: if danger can be avoided later, it poses no threat now. We show how a violation of this assumption --- a pessimistic, false belief that later avoidance will be unsuccessful --- leads to a characteristic propagation of fear and avoidance to situations far antecedent of threat. This single deviation can explain a surprising range of features of anxious behavior, including exaggerated threat appraisals, fear generalization, and persistent avoidance. Simulations of the model reproduce laboratory demonstrations of abnormal decision making in anxiety, including in situations of approach-avoid conflict and planning to avoid losses. The model also ties together a number of other seemingly disjoint issues in anxious disorders. For instance, learning under the pessimistic bias captures a hypothesis about the role of anxiety in the later development of depression. The bias itself offers a new formalization of classic insights from the psychiatric literature about the central role of maladaptive beliefs about control and self-efficacy in anxiety. This perspective is importantly different from previous computational accounts of beliefs about control in mood disorders, which neglected the sequential aspects of choice.
\end{abstract}

@@

\acknowledgements{This research was supported in part by NIH grant MH10917, part of the CRCNS program.}

@@

The puzzle of these symptoms is particularly sharp from a decision theoretic perspective \citep{huys2015}. Here we isolate and address a  decision theoretic paradox that, we argue, clarifies the core psychopathology: The generalization of fear and avoidance to situations far antecedent of threat violates the basic logic of evaluation over sequential trajectories of action. This is because avoidance is by nature protective: the ability to avoid in future means I need not also do so now. For instance, cars endanger pedestrians but can be reliably avoided by following traffic signals; given that, staying indoors offers no additional protection from accidents. This is an instance of a general property of evaluating actions in sequential tasks: such evaluation turns fundamentally on assumptions about subsequent events, including the agent's own subsequent choices. Typically, it is appropriate to assume that the agent will also make good choices down the line.

This line of reasoning, then, hints that a fundamental deficit in anxiety disorders may relate to this assumption, which otherwise should preclude the spread of avoidance. Indeed, anxiety disorders are associated with pessimistic beliefs about the future \citep{ClarkBeck2011}. For instance, anxious individuals are more likely to endorse beliefs like that they are likely to encounter danger in the future and, notably, that they will be unable to mitigate it when they do. Here we develop this idea into a model of evaluation under pessimistic assumptions about future choices. We show that this single, localized deviation can explain a surprising range of features of anxious behavior including exaggerated threat appraisal, fear generalization, and persistent avoidance. This model also offers a new formalization of classic insights from the psychiatric literature about the central role of beliefs about control and self-efficacy in anxiety. Our perspective is importantly different from previous computational accounts of beliefs about control in mood disorders (e.g. \cite{HuysDayan2009}), which neglected the sequential aspects of choice.

@@

Consider an agent who instead has pessimistic expectations about their future actions. Such returns can be adaptive to ensure robustness such as under adversarial scenarios \cite{Garcia2015}; here we propose unrealistically pessimistic assumptions as a root cause of many anxious symptoms. We adopt the $\beta$-pessimistic value function from \cite{Gaskett2003}, to define
\begin{equation*}
Q^w(s,a) = r(s,a) + \gamma \sum_{s'} p(s' \mid s,a) \left( w \max_{a'} Q^w(s',a') + (1 - w) \min_{a'} Q^w(s',a') \right)
\end{equation*}
Here, the parameter $w$ controls the degree of pessimism. An optimistic agent ($w = 1$) expects in the future to act fully in accordance with its preferences whereas a pessimistic agent ($w = 0$) expects to act contrary to its preferences. Fig.~1c,d illustrates the the consequences of different levels of pessimism for valuation in the example gridworld; here, the expectation that the agent may fail to avoid in future causes the value of threat to permeate the state space.

@@

This simple simulation --- reflecting a localized violation of a core decision theoretic assumption --- already echoes several core symptoms of anxiety disorders. The pessimistic agents in Fig~1c,d exhibit exaggerated threat appraisals (otherwise neutral states signal danger); generalization of fear (threat value spreads across the gridworld); and persistent avoidance (the agent chooses paths which maintain increasing distance from threat).

@@

\textbf{Aversive pruning:} Another empirical phenomenon associated with anxiety is ``aversive pruning'' in planning \cite{Huys2012, Lally2017}. This refers to the idea that when evaluating future trajectories in a sequential task like chess, people must selectively consider certain paths and neglect others. One proposal for how people accomplish this is aversive pruning \citep{Huys2012}: choice sequences involving large losses are discarded from further evaluation. For instance, although the optimal choice in the decision tree from Fig.~3a is to weather an initial large loss in order to reap the large gain that follows, people tend to disfavor this path suggesting they neglect the later gain. The degree of such pruning correlates with depressive \cite{Huys2012} and anxiety \cite{Lally2017} symptoms. Our model predicts this result (Fig.~3b), though for a somewhat different reason: in our model, pessimistic (anxious) agents neglect large gains deeper in the tree, not because they fail to consider them, but because with increasing anxiety they increasingly predict the possibility of choosing incorrectly afterward, and failing to recoup the loss. Future research could use slight variants in the decision trees to tease apart these different interpretations.

@@

A second point about learning in this model is that, because the pessimistic return ultimately spreads biased value estimates to states and actions far distant from the primary danger (e.g., Fig.~1d), it would accordingly also take a great many steps of iterative learning to correct all these misattributions, even given direct experience that the danger itself has passed. This may offer at least a partial answer to a classic puzzle in pathological avoidance, i.e. why it is so resistant to extinction \cite{moutoussis2018}.

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/04_tree_rldm.png}}%
  }
  \par \textbf{Figure 3:} The decision tree environment from \cite{Huys2012} and \cite{Lally2017} (A). An optimistic agent ($w=1$) prefers the optimal loss-minimizing policy through the initial large loss (B). A pessimistic agent ($w=0.5$) comes to prefer the branch without the large loss so as to avoid being unable to recoup the large initial loss. One-step rewards are presented in each state; the net value $Q$ on each path is shown numerically.
\end{figure}

@@

One important ambiguity is that we have formalized pessimistic assumptions in terms of beliefs about the agent's own future actions: expecting failure, for instance, to avoid in future. Since Eq.~1 is also computed in expectation over the anticipated future environmental dynamics $p(s' \mid s,a)$, pessimism can alternatively be encoded in this distribution (e.g., a belief that the world's response to one's choices is unpredictable or adversarial). Because the Bellman equation averages out both distributions,  and because an unpredictable environment also reduces the efficacy of avoidance, either formulation can produce ultimately similar results. However from the perspective of cognitive theories of anxiety, these represent quite different maladaptive beliefs, which may be relevant, for instance, in guiding treatment.
